---
title: "Linearizing a multidimensional distribution"
author: "Karsten W."
date: "2022-12-10"
output: html_document
vignette: >
  %\VignetteIndexEntry{Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Oftentimes, the probability distribution we wish to estimate with the maximum entropy principle is multidimensional. This distribution is represented in the algorithms as one numerical vector. One challenge -- or at least annoying and error-prone task -- is to map a subset of the probability distribution to this vector.

The `fd.maxent` package offers a few helper functions to make this easier. This vignette explains the ideas and the usage of these functions. The "europawahl 2019" problem is used as example. While there is a separate vignette showing how to apply the maximum entropy solver, this vignette only looks at the linearizing topic. 

## Multidimensional Probability Distribution Example: Voting

Assume we would like to investigate voter's behaviour according to some sociodemographic characteristics. (Please note I'm not a political scientist, this example is purely for showing the concept of the functions.) 

Then we would be interested in a probability distribution that is the Cartesian product of
the voter's gender, voter's education and the party voted for. Within this package, you could represent this distribution thus:

```{r defs, echo=TRUE}
library(fd.maxent)
gender <- set("female", "male")
edu <- set("Hochschule", "Abitur", "mittlere Reife", "Hauptschule")
voted_for <- set("CDU/CSU", "Gr\u00fcne", "SPD", "AfD", "Other", "Invalid", "Not voted")
prob <- variab(vote=voted_for, mf=gender, edu=edu)
```

The function `set` is defined in the `fd.maxent` package and defines one dimension of the probabilty distribution. Only `character` arguments are allowed here.

The `edu` (education) and `voted_for` (parties) dimensions are specific to the German voting system.

The joint distribution is defined with the `variab` function. Since it is the Cartesian product of its dimensions, it has `2x4x7==56` cells. This can be verified via `length`:

```{r len, echo=TRUE}
length(prob)
```

So a constraint matrix for a maximum entropy problem would have `56` columns. 

## Accessing one Cell

Assume we know the exact non-voting percentage for female voters with a "Hauptschule" eduction. How could we code this knowledge as a constraint in the maximum entropy problem? We would need which index, 1..56, refers to this specific cell.

This is where the `index` function from the `fd.maxent` package comes in:

```{r cell, echo=TRUE}
index(prob, vote="Not voted", mf="female", edu="Hauptschule")
```

So index `49` refers to this specific cell. We can use this information in `mep_set_constraint`.

## Indices for a Margin

Let's take another example. Assume we know the exact non-voting percentage for all males, and have no information how this percentage is distributed across the education dimension. Then we could include a constraint in the maximum entropy problem that uses the sum of a cells with male voters and assign as right-hand side the known percentage.

Again, the `index` can help with that:

```{r margin, echo=TRUE}
index(prob, vote="Not voted", mf="male")
```

So, the four cells `14 28 42 56` represent the male non-voters. It must be four because there are four characteristics for education.

## De-Linearizing

Assume we have a result from the solver and want to interpret this result in terms of the probability distribution. Here comes the `vec2df` function into play:

```{r vec2df, echo=TRUE}
x <- runif(length(prob)) # fake data instead of solver output
vec2df(prob, x) |> knitr::kable()
```


